Predicting the (non-cryptographic) PCG Pseudo-Random Number Generator In Practice

### 2020-05-30 article/main.tex
intro ferrenberg

\todo{Mentionner Ferrenberg 92}

The scientific computing community also realized that the need for fast
\emph{parallel} random number generation could be satisfied by the use of block
ciphers in counter mode~\cite{Salmon11}. The need for speed then leads to the
use of weakened cryptographic primitives.

In most cases, it is easy to see that a non-cryptographic PRG does not meet the
cryptographic notion of pseudo-randomness, and there are few exceptions. In this
paper, we study the \textsf{PCG} family of non-cryptographic pseudo-random
generators~\cite{melissapaper,melissaweb}.

@@

\paragraph{Contribution.} We describe an algorithm that reconstructs the full
internal state of the strongest member of the \textsf{PCG} family. This allows
to predict the pseudo-random stream deterministically and clock the generator
backwards. The original seeds can also easily be reconstructed. The algorithm is
practical and we have executed it in practice. It follows that predicting the
output of the \textsf{PCG} should not be considered challenging anymore.

@@

When the increment is unknown, things are a bit more complicated. This is the
default situation in \textsf{NumPy}, where both the state and the increment are
initialised using an external source of entropy. In this case, our prediction
algorithm requires 3072 bits of pseudo-random stream ; it guesses between 51 and
55 bits, then for each guess it solves and instance of CVP in dimension 4 (using
about 75 arithmetic operations). This recovers 64 more bits of information about
the difference between two successive states, and this is enough to filter the
bad guesses. This information can then be used in a subsequent and comparably
inexpensive phase to recover the entire internal state. On average, the whole
process requires a bit less than 20 000 CPU hours to complete.

@@

\paragraph{Related Work.} \todo{Mentionner Knuth et les LCG + Frieze + Boyar + Stern...}

@@

The designer of \textsf{PCG} was kind enough to send us challenge inputs. She
generated random seeds and provided us with the first outputs of the
pseudo-random generator. We were able to reconstruct the seed with an extremely
high confidence level, because they re-generate the same outputs. We emailed
back the seeds and received confirmation that they were indeed correct.

>>>>

Failures in cryptographic pseudo-random generators have catastrophic security
implications. Let us mention for instance the well-known problem in Debian
Linux, where a bug in the OpenSSL package led to insufficient entropy gathering
and to practical attacks on the SSH and SSL protocols (the only remaining source
of entropy comes from the PID of the process, i.e. 16 bits or less of effective
entropy)~\cite{YilekRSES09}.

However, problems in non-cryptographic random number generators can also have
problematic consequences (barring the obvious case where they are used in lieu
of their cryptographic counterparts). When they are used in scientific computing
for Monte-Carlo simulations, their defects have the potential to alter the
results of the simulations. Ferrenberg et al.~\cite{Ferrenberg92} ran a
classical Ferromagnetism Ising model Monte-Carlo simulation, in a special case
where the exact results could be computed analytically, and compared the results
of the simulation with the ``true'' answer. They used several conventional
pseudo-random generators: a 32-bit linear congruential generator, two LFSRs,
various combinations thereof, etc. They observed that changing the source of
random numbers significantly altered the outcome of the physical
simulation. Different generators produced different biases: in particular a
given LFSRs yielded energy levels that were systematically too low and critical
temperatures that were always too high, while another kind of generator yielded
the opposite.

The scientific computing community also realized that the need for fast
\emph{parallel} random number generation could be satisfied by the use of block
ciphers in counter mode~\cite{Salmon11}. The need for speed then leads to the
use of weakened cryptographic primitives (round-reduced AES or custom and
presumably weak block-ciphers)

In most cases, it is fairly easy to see that a specific non-cryptographic PRG
does not meet the cryptographic notion of pseudo-randomness, and there are few
exceptions. Most are fairly easy to predict, meaning that after having observed
a prefix of the output, it is easy to produce the next ``pseudo-random''
bits.

In this paper, we study the \textsf{PCG} family of non-cryptographic
pseudo-random generators~\cite{melissapaper,melissaweb}. Its designer claimed
that predicting it ought to be ``challenging'', even though cryptographic
strength is not claimed.

@@

\paragraph{Contribution.} We describe an algorithm that reconstructs the full
internal state of the strongest member of the \textsf{PCG} family. This allows
to predict the pseudo-random stream deterministically and clock the generator
backwards. The original seeds can also easily be reconstructed. The state
reconstruction algorithm is practical and we have executed it in practice. It
follows that predicting the output of the \textsf{PCG} should not be considered
challenging anymore.

While the \textsf{PCG} pseudo-random generator is not meant as a cryptographic
primitive, obtaining an actual prediction algorithm requires the use of (fairly
conventional) cryptanalytic techniques.

@@

When the increment is unknown, things are a bit more complicated. This is the
default situation in \textsf{NumPy}, where both the state and the increment are
initialised using an external source of entropy. In this case, our prediction
algorithm requires 3072 bits of pseudo-random stream ; it guesses between 51 and
55 bits, then for each guess it solves an instance of CVP in dimension 4 (using
about 75 arithmetic operations). This recovers 64 more bits of information about
the difference between two successive states, and this is enough to filter the
bad guesses. This information can then be used in a subsequent and comparably
inexpensive phase to recover the entire internal state. On average, the whole
process requires a bit less than 20 000 CPU hours to complete.

@@

\paragraph{Related Work.} Knuth considered the problem of recovering the
internal state of a truncated linear congruential generator
$x_{i+1} = ax_i + c \bmod 2^k$ when $a$ and $c$ are unknown~\cite{Knuth85}; he
gave an algorithm exponential in the number of truncated bits.

\todo{Mentionner Knuth et les LCG + Frieze + Boyar + Stern...}

@@

The designer of \textsf{PCG} was kind enough to send us two sets challenge
inputs: one with the default (known) increment and one with a random secret
increment. She generated random seeds and provided us with the first outputs of
the pseudo-random generator. We were able to reconstruct the seed with an
extremely high confidence level, because they re-generate the same outputs. We
emailed back the seeds and received confirmation that they were indeed correct.

We have therefore successfully taken the challenge of predicting the output of
the \textsf{PCG64} generator.
